preciseTAD: A machine learning framework for precise TAD boundary prediction at base level resolution

### 2020-08-15 Manuscript/manuscript.Rmd
Grammarly fixes

**Results** Using different feature engineering and resampling strategies, we found spatial relationships (distance) between boundaries and annotations have the most predictive power. Transcription factor binding sites outperformed other genome annotation types. Random under-sampling significantly improved model performance. Boundaries predicted by preciseTAD were more enriched for CTCF, RAD21, SMC3, and ZNF143 sites and more conserved across cell lines, highlighting their higher biological significance. The model trained in one cell line perform well in predicting boundaries in another cell line.

@@

The advent of chromosome conformation capture (3C) sequencing technologies, and its successor Hi-C, have revealed a hierarchy of 3-dimensional (3D) spatial constructs within the human genome such as chromatin loops [@rao20143d], Topologically Associated Domains (TADs) [@nora2012spatial;@dixon2012topological], and A/B compartments [@lieberman2009comprehensive]. At kilobase scale, gene promoters are often connected via distal regulatory elements, such as enhancers, through chromatin loops, thereby regulating gene expression [@Lupianez:2015aa; @Franke:2016aa]. Further, at the megabase scale, TADs represent regions on the linear genome, that are highly self-interacting. Emerging evidence has linked TADs to critical roles in cell dynamics and cell differentiation. Studies have shown that TADs themselves are highly conserved across species and cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa]. Likewise, disruption of boundaries demarcating TADs and loops promotes cancer [@hnisz2016activation; @taberlay2016three] and other disorders [@lupianez2016breaking;@Sun:2018aa;@Meaburn:2007aa]. Therefore, precisely identifying the location of TAD and chromatin loop boundaries remains a top priority in our goal to fully understand the functionality of the human genome.

Several methods have been proposed to identify genomic coordinates of TADs (reviewed in [@Zufferey:2018aa; @forcato2017comparison]), and chromatin loops [@rao20143d; @Ay:2014aa; @Heinz:2010aa; @salameh2019supervised].  However, a key limitation among most of these TAD- and loop callers is that they are heavily reliant on Hi-C data *resolution*. Resolution refers to the sizes of genomic regions (bins) used to segment the linear genome [@li2019comparison]. Lower resolutions contribute to larger bin sizes, which lead to uncertainty in the location of boundaries. Due to high sequencing costs, most available Hi-C datasets have relatively low resolution, ranging from 25 kb to 100 kb. Because increasing the resolution of Hi-C data requires a quadratic increase in the total sequencing depth, obtaining high resolution remains difficult [@schmitt2016genome]. While these low resolution Hi-C datasets can be used to probe higher order genomic patterns such as A/B compartments, more refined resolutions are necessary to accurately define TAD and loop boundaries [@fudenberg2016formation].

@@

We present *preciseTAD*, an R package implementing an optimally tuned machine learning framework for precise identification of domain boundaries using genome annotation data. Our method utilizes the random forest (RF) algorithm trained on chromatin segmentation (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS) data. We introduce a systematic pipeline for building optimal boundary prediction classifier. We found that spatial associations (distance) between boundaries and annotations perform best, transcription factor binding sites further improve prediction accuracy, and a simple random undersampling technique effectively addresses the negative effect of class imbalance. Translated from Hi-C data resolution level to base level (annotating each base and predicting its boundary probability), *preciseTAD* employs density-based clustering (DBSCAN) and scalable partitioning around medoids (CLARA) on prediction probabilities to select genome annotation-guided boundary coordinates. This approach circumvents resolution restrictions of Hi-C data, allowing for the precise detection of biologically meaningful boundaries. We demonstrate that predictions made by *preciseTAD* are more enriched for known molecular drivers of 3D chromatin including CTCF, RAD21, SMC3, and ZNF143. Further, we show that *preciseTAD* boundaries are more conserved across cell lines, and the model trained in one cell line can predict boundaries in another cell line. This improved precision in domain boundary location can provide insight into the association between genomic regulators and the 3D genome organization. The *preciseTAD* R package is freely available on GitHub (https://github.com/dozmorovlab/preciseTAD) and on Bioconductor (submitted).

@@

Publicly available TAD and loop boundaries called by Arrowhead [@durand2016juicebox] and Peakachu [@salameh2019supervised] tools, respectively, were used for training. Arrowhead TAD boundaries were called from primary Hi-C data for the GM12878 and K562 cell lines [@rao20143d]. The contact matrices were constructed from all read pairs that uniquely aligned to the genome (MAPQ>0). The default parameters for Arrowhead were used when extracting the autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries at 5 kb, 10 kb, 25 kb, 50 kb, and 100 kb resolutions (Additional File 1). Peakachu chromatin loop boundaries called at 10 kb for the GM12878 and K562 cell lines were downloaded from the Yue lab web site (Supplementary Table 1). Unique boundaries were considered as the midpoints within the coordinate of each chromatin loop anchor. Chromosome 9 was excluded from all downstream analyses due to the sparsity of contact matrices at 5 kb and 10 kb resolutions for the K562 cell line. Cell-line-specific genomic annotations (BroadHMM chromatin segmentations (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS)) were obtained from the UCSC Genome Browser Database [@Tyner:2017aa] (Supplementary Table 2).

@@

To create shifted binning, the first shifted bin was set to start at half of the resolution *r*, and continued in intervals of length *r* until the end of the chromosome (*mod r* + *r/2*). The shifted bins, referred hereafted as bins for simplicity, were then defined as boundary-containing regions (*Y = 1*) if they contained a TAD (or loop) boundary, and non-boundary regions (*Y = 0*) otherwise, thus establishing the binary response vector (**Y**) used for classification (Figure 1A).

@@

To assess the impact of class imbalance (CI), defined as the proportion of boundary regions to non-boundary regions, we evaluated three re-sampling techniques. These include *Random Over-Sampling (ROS)*, *Random Under-Sampling (RUS)*, and *Synthetic Minority Over-Sampling Technique (SMOTE)*. For ROS, the minority class was sampled with replacement to obtain the same number of data points in the majority class. For RUS, the majority class was sampled without replacement to obtain the same number of data points in the minority class. For SMOTE, under-sampling was performed without replacement from the majority class, while over-sampling was performed by creating new synthetic observations using the $k=5$ minority class nearest neighbors [@chawla2002smote] (implemented in the `DMwR` v.0.4.1 R package). We restricted the SMOTE algorithm to 100% over-sampling and 200% under-sampling, which creates perfectly balanced classes.

@@

Random forest (RF) classification models (the `caret` v.6.0 R package [@kuhn2012caret]) were built to compare model performances between combinations of data resolutions, feature engineering procedures, and re-sampling techniques. A *holdout chromosome* technique was used for model evaluation on each autosomal chromosome. The $i^{th}$ holdout chromosome was identified and a data matrix, $A_{N \times (p+1)}$, was constructed by combining the binned genome from the remaining chromosomes ($1,2,\cdots, i-1,i+1,\cdots,21,22$), where $N=[n_{1} n_{2} \cdots n_{21} n_{22}]'$ and $n_{k}$ is the length of chromosome *k* after being binned into non-overlapping regions of resolution *r*, such that $k \ne i$. $p$, the number of annotations, together with the response vector **Y**, defined the column-wise dimension of the matrix $A$. Re-sampling was then performed on $A$ and an RF classifier was trained using 3-fold cross-validation to tune for the number of annotations to consider at each node (*mtry*). The number of trees (*ntree*) that were aggregated for each RF model was set to 500. The minimum number of observations per root node (*nodesize*) was set to .01% of the rows in the data. The binned data for the holdout chromosome $i$ was reserved for testing. Models were evaluated using Balanced Accuracy (BA), defined as the average of sensitivity and specificity:

@@

where TP refers to the number of bins correctly identified as containing a boundary (true positives), FP refers to the number of bins incorrectly identified as containing a boundary (false positives), TN refers to the number of bins correctly identified as not containing a boundary (true negatives), and FN refers to the number of bins incorrectly identified as not containing a boundary (false negatives). Each of these quantities are obtained from the confusion matrix created by validating the model on the test data. The process was repeated for each $i^{th}$ holdout chromosome and performances were aggregated using mean and standard deviation.

@@

Many genomic annotations, notably architectural proteins, tend to exhibit an extensive pattern of colocalization (correlation) [@van2014insulator]. To suitably reduce the predictor space and improve computational efficiency, while maintaining optimal performance, we implemented recursive feature elimination (RFE). In doing so, we estimated the near optimal number of necessary features to include in the best fitting RF model. The best performing subset was identified, ranging from 2 to the maximum number of features incremented by the power of 2. We then aggregated the predictive importance of the union of the optimal set of features across chromosomes using the mean decrease in node impurity among permuted features in out-of-bag samples to determine the most common and top ranked annotations for predicting boundary regions.

@@

When evaluating performance across cell lines, a random forest model was built using ground truth boundaries and genomic annotations from one cell line such that the data matrix was constructed using genomic bins from $n-1$ chromosomes. The test data used to evaluate performance was constructed from the binned data of the $i^{th}$ holdout chromosome with genomic annotations from the other cell line. Thus, the models were trained on cell line specific data and tested on the other cell line. This process was repeated for each holdout chromosome. To evaluate performance, we constructed receiver operating characteristic (ROC) curves composed of the average sensitivies and specificities, and reported the area under the curve (AUC).

@@

To investigate whether we could alleviate resolution limitations of conventional domain calling tools, we developed *preciseTAD*, an algorithm that leverages a random forest model in conjunction with density-based and partitioning techniques to predict boundaries at base resolution. A random forest model was built on the optimal combination of predictor type (predictor space), re-sampling technique (class balancing correction), and top ranked annotations (most informative genomic elements) to maximally accurately predict boundaries at the _Hi-C resolution level_.

@@

Next, we assessed the biological significance of our predicted boundaries by their association with several known molecular drivers of 3D chromatin architecture including CTCF, RAD21, SMC3, and ZNF143 ChIP-Seq peaks. To do so, we used the *computeMatrix* and *plotHeatmap* tools offered by deepTools (version 2.0) [@ramirez2016deeptools] via integration with Galaxy (version 3.3.2.0.0) [@afgan2018galaxy]. Additionally, we compared the median $log_2$ genomic distances between TAD boundaries and the same top predictive ChIP-seq annotations using Wilcoxon Rank-Sum tests. Furthermore, we compared the overlap between predicted and called boundaries in GM12878 and K562 cell lines. Boundaries were first flanked by resolution, *r*, and overlaps were visualized using Venn diagrams from the `Vennerable` R package (version 3.1.0). Overlaps were further quantified using the Jaccard index defined as

@@

## Developing a ML framework for optimal TAD boundary prediction

We developed a machine learning (ML) framework aimed at determining the optimal set of data level characteristics necessary for precise prediction of boundaries of Topologically Associated Domains (TADs) and chromatin loops, collectively referred to as domain boundaries. We chose the random forest (RF) algorithm as our binary classification tool. The reason for it is two-fold: (1) to devise a tunable prediction rule in a supervised learning framework that is both robust to overfitting and able to handle multiple correlated predictors, and (2) to allow for an interpretable ranking of predictors [@boulesteix2012overview]. We use published Arrowhead-called TAD boundaries [@durand2016juicer] and Peakachu-predicted loop boundaries [@salameh2019supervised] as ground truth. Data from GM12878 and K565 cell lines at 5-100kb resolution (Arrowhead) and 10kb resolution (Peakachu) were used (Supplementary Table 1). We next developed *preciseTAD*, an R package which leverages an optimally built RF to precisely predict genomic coordinates of TAD and loop boundaries at base resolution from cell type-specific genome annotation data.

Boundary regions were defined as genomic bins containing a called boundary ($Y = 1$), while non-boundary regions were defined as bins that did not contain a called boundary ($Y = 0$) (Figure 1A, see Methods). The total number of called TADs, their unique boundaries, and the number of genomic bins expectedly decreased with the decreased resolution of Hi-C data (Table 1, Supplementary Table 3). The number of non-boundary regions highly outnumbered boundary regions. Such disproportional presence of examples in one class is known as a "class imbalance" problem that negatively affects predictive modeling [@prati2011survey; @wei2013role]. To address class imbalance, we evaluated the effect of three re-sampling techniques. *Random over-sampling* (ROS) was defined as sampling with replacement from the minority class (boundary regions). *Random under-sampling* (RUS) was defined as sampling with replacement from the majority cluss (non-boundary regions). Lastly, we tested *Synthetic minority over-sampling technique* (SMOTE), which is a combination of both random over- and under-sampling to create balanced classes [@chawla2002smote] (see Methods).

A total of 77 cell-line-specific genomic annotations were used to build the predictor space. These included BroadHMM chromatin segmentations, histone modifications (HM), and transcription factor binding sites (TFBS) (Supplementary Table 2). Four feature engineering procedures were developed to quantify the association between genomic annotations and bins (Figure 1B). These included signal strength association (Signal), direct (OC), proportional (OP), and spatial ($log_2 + 1$ Distance) relationships (see Methods, Supplementary Figure 1).

![**Resolution-specific data construction and feature engineering used for downstream random forest modeling.** (A) The linear genome was binned into non-overlapping resolution-specific intervals (see Methods). The response vector **Y** was defined as 1/0 if a genomic bin overlapped/did not overlap with a TAD boundary, respectively. (B) Four types of associations between bins and genomic annotations (green shapes) were considered to build the predictor space including Average Peak Signal (Signal), Overlap Counts (OC), Overlap Percent (OP), and $log_{2}$ distance (Distance).](Figures/figure1.png)

In total, we considered combinations of data from two cell lines *L = {GM12878, K562}*, five resolution *R = {5 kb, 10 kb, 25 kb, 50 kb, 100 kb}*, four types of predictor space *P = {Signal, OC, OP, Distance}*, and three re-sampling technique *S = {None, RUS, ROS, SMOTE}* (Figure 2). Once the model inputs were established, a random forest classifier was trained on $n-1$ autosomal chromosomes, while reserving the $i^{th}$ chromosome for testing. Three-fold cross-validation was used to tune the *mtry* hyperparameter, while *ntree* and *nodesize* were fixed to default values of 500 and 1 respectively. Model performance was evaluated by aggregating the mean balanced accuracy (BA) across each holdout chromosome (see Methods). These strategies allowed us to select the best performing model in an unbiased manner.

![**A diagram of the machine learning framework used to build TAD/loop boundary region prediction models.** Step 1 defines the dimensions of the data matrix $A_{N \times (p+1)}$ used for training and testing by establishing a resolution, $r$, at which to bin the linear genome into $N$ genomic bins for each chromosome, except a holdout chromosome *i*. The response vector **Y** is defined as a boundary region ($Y=1$) if it overlaps with a genomic bin (else $Y=0$) A feature engineering procedure is selected to define the feature space of $A$ composed of $p$ genomic annotations. Step 2 is to reserve the predictor-response matrix for the holdout chromosome *i* as the test data. Step 3 is to apply a re-sampling technique to the training data to address class imbalance. Step 4 is to train the random forest model and perform 3-fold cross-validation to tune for the mtry parameter. Finally, step 5 is to validate the model on the separate test data composed of the binned data from the holdout chromosome *i* and to evaluate model performance using balanced accuracy (BA).](Figures/figure2.png)

@@

Expectedly, when using data with class imbalance present, that is, no re-sampling, the models exhibited low balanced accuracies, with very little variability among different resolutions (Figure 3). Similarly, poor performances were found when using ROS. However, RUS and SMOTE re-sampling led to drastic improvement in performance, especially at higher resolutions. We found that RUS marginally outperformed SMOTE as the most optimal class balancing technique for all resolutions and predictor types when predicting TAD boundary regions.

Additionally, we found that using a distance-type predictor space yielded substantially higher balanced accuracies when compared to the peak signal stength, overlap count, and overlap percent predictor types. As with class balancing techniques, this improvement was less evident at lower resolutions with results consistent for K562 (Supplementary Figure 2). Random forest models built on Peakachu-defined loop boundaries with SMOTE resampling only marginally outperformed RUS for both GM12878 and K562 cell lines (Supplementary Figure 3). Furthermore, 5 kb resolution genomic bins led to optimal prediction for both TAD and loop boundaries, on both cell lines. Our results indicate that random under-sampling, distance-type predictors, and high-resolution Hi-C data provide the optimal set of data level characteristics for both TAD and chromatin loop boundary prediction.

![**Distance-type predictors combined with random under-sampling at 5 kb resolution genomic bins yielded optimal TAD boundary region predictive performance for GM12878 Arrowhead TAD boundaries.** Averaged balanced accuracies are compared across resolution, within each predictor-type: (A) Signal, (B) OC, (C) OP, and (D) Distance, and across re-sampling techniques: no re-sampling (None; red), random over-sampling (ROS; green), random under-sampling (RUS; blue), and synthetic minority over-sampling (SMOTE; purple). Error bars indicate standard deviation away from the mean performance across each holdout chromosome used for testing.](Figures/figure3.png)

@@

We hypothesized that the type of genomic annotations may also affect predictive performance. Using the established optimal settings (RUS, Distance, 5 kb bins), we built separate random forest models using BroadHMM chromatin segmentations, histone modifications (HM), and transcription factor binding sites (TFBS). We found that models built on TFBSs outperformed other annotation-specific models, with results consistent for loop boundaries, on both cell lines (Figure 4A; Supplementary Figures 4A). These results suggest that TFBS are the primary drivers of TAD and loop boundary formation in both GM12878 and K562.

## Predictive importances confirmed biological role of CTCF, RAD21, SMC3, and ZNF143 for boundary formation

>>>>

**Results** Using different feature engineering and resampling strategies, we found spatial relationships (distance) between boundaries and annotations have the most predictive power. Transcription factor binding sites outperformed other genome annotation types. Random under-sampling significantly improved model performance. Boundaries predicted by preciseTAD were more enriched for CTCF, RAD21, SMC3, and ZNF143 sites and more conserved across cell lines, highlighting their higher biological significance. The model trained in one cell line performs well in predicting boundaries in another cell line.

@@

The advent of chromosome conformation capture (3C) sequencing technologies, and its successor Hi-C, have revealed a hierarchy of 3-dimensional (3D) spatial constructs within the human genome such as chromatin loops [@rao20143d], Topologically Associated Domains (TADs) [@nora2012spatial;@dixon2012topological], and A/B compartments [@lieberman2009comprehensive]. At the kilobase scale, gene promoters are often connected via distal regulatory elements, such as enhancers, through chromatin loops, thereby regulating gene expression [@Lupianez:2015aa; @Franke:2016aa]. Further, at the megabase scale, TADs represent regions on the linear genome, that are highly self-interacting. Emerging evidence has linked TADs to critical roles in cell dynamics and cell differentiation. Studies have shown that TADs themselves are highly conserved across species and cell lines [@dixon2012topological; @krefting2018evolutionary; @nora2012spatial; @dixon2015chromatin; @rao20143d; @Schmitt2016aa]. Likewise, disruption of boundaries demarcating TADs and loops promotes cancer [@hnisz2016activation; @taberlay2016three] and other disorders [@lupianez2016breaking;@Sun:2018aa;@Meaburn:2007aa]. Therefore, precisely identifying the location of TAD and chromatin loop boundaries remains a top priority in our goal to fully understand the functionality of the human genome.

Several methods have been proposed to identify genomic coordinates of TADs (reviewed in [@Zufferey:2018aa; @forcato2017comparison]), and chromatin loops [@rao20143d; @Ay:2014aa; @Heinz:2010aa; @salameh2019supervised].  However, a key limitation among most of these TAD- and loop callers is that they rely heavily on Hi-C data *resolution*. Resolution refers to the sizes of genomic regions (bins) used to segment the linear genome [@li2019comparison]. Lower resolutions contribute to larger bin sizes, which lead to uncertainty in the location of boundaries. Due to high sequencing costs, most available Hi-C datasets have relatively low resolution, ranging from 25 kb to 100 kb. Because increasing the resolution of Hi-C data requires a quadratic increase in the total sequencing depth, obtaining high resolution remains difficult [@schmitt2016genome]. While these low-resolution Hi-C datasets can be used to probe higher-order genomic patterns such as A/B compartments, more refined resolutions are necessary to accurately define TAD and loop boundaries [@fudenberg2016formation].

@@

We present *preciseTAD*, an R package implementing an optimally tuned machine learning framework for precise identification of domain boundaries using genome annotation data. Our method utilizes the random forest (RF) algorithm trained on chromatin segmentation (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS) data. We introduce a systematic pipeline for building optimal boundary prediction classifier. We found that spatial associations (distance) between boundaries and annotations perform best, transcription factor binding sites further improve prediction accuracy, and a simple random undersampling technique effectively addresses the negative effect of class imbalance. Translated from Hi-C data resolution level to base level (annotating each base and predicting its boundary probability), *preciseTAD* employs density-based clustering (DBSCAN) and scalable partitioning around medoids (CLARA) on prediction probabilities to select genome annotation-guided boundary coordinates. This approach circumvents resolution restrictions of Hi-C data, allowing for the precise detection of biologically meaningful boundaries. We demonstrate that *preciseTAD* predictions are more enriched for known molecular drivers of 3D chromatin, including CTCF, RAD21, SMC3, and ZNF143. Further, we show that *preciseTAD* boundaries are more conserved across cell lines, and the model trained in one cell line can predict boundaries in another cell line. This improved precision in the domain boundary location can provide insight into the association between genomic regulators and the 3D genome organization. The *preciseTAD* R package is freely available on GitHub (https://github.com/dozmorovlab/preciseTAD) and on Bioconductor (submitted).

@@

Publicly available TAD and loop boundaries called by Arrowhead [@durand2016juicebox] and Peakachu [@salameh2019supervised] tools, respectively, were used for training. Arrowhead TAD boundaries were called from primary Hi-C data for the GM12878 and K562 cell lines [@rao20143d]. The contact matrices were constructed from all read pairs that uniquely aligned to the genome (MAPQ>0). The default parameters for Arrowhead were used when extracting the autosomal genomic coordinates (GRCh37/hg19 human genome assembly) of TAD boundaries at 5 kb, 10 kb, 25 kb, 50 kb, and 100 kb resolutions (Additional File 1). Peakachu chromatin loop boundaries called at 10 kb for the GM12878 and K562 cell lines were downloaded from the Yue lab website (Supplementary Table 1). Unique boundaries were considered as the midpoints within the coordinate of each chromatin loop anchor. Chromosome 9 was excluded from all downstream analyses due to the sparsity of contact matrices at 5 kb and 10 kb resolutions for the K562 cell line. Cell-line-specific genomic annotations (BroadHMM chromatin segmentations (BroadHMM), histone modifications (HM), and transcription factor binding sites (TFBS)) were obtained from the UCSC Genome Browser Database [@Tyner:2017aa] (Supplementary Table 2).

@@

To create shifted binning, the first shifted bin was set to start at half of the resolution *r* and continued in intervals of length *r* until the end of the chromosome (*mod r* + *r/2*). The shifted bins, referred hereafter as bins for simplicity, were then defined as boundary-containing regions (*Y = 1*) if they contained a TAD (or loop) boundary, and non-boundary regions (*Y = 0*) otherwise, thus establishing the binary response vector (**Y**) used for classification (Figure 1A).

@@

To assess the impact of class imbalance (CI), defined as the proportion of boundary regions to non-boundary regions, we evaluated three resampling techniques. These include *Random Over-Sampling (ROS)*, *Random Under-Sampling (RUS)*, and *Synthetic Minority Over-Sampling Technique (SMOTE)*. For ROS, the minority class was sampled with replacement to obtain the same number of data points in the majority class. For RUS, the majority class was sampled without replacement to obtain the same number of data points in the minority class. For SMOTE, under-sampling was performed without replacement from the majority class, while over-sampling was performed by creating new synthetic observations using the $k=5$ minority class nearest neighbors [@chawla2002smote] (implemented in the `DMwR` v.0.4.1 R package). We restricted the SMOTE algorithm to 100% over-sampling and 200% under-sampling, which creates perfectly balanced classes.

@@

Random forest (RF) classification models (the `caret` v.6.0 R package [@kuhn2012caret]) were built to compare model performances between combinations of data resolutions, feature engineering procedures, and resampling techniques. A *holdout chromosome* technique was used for model evaluation on each autosomal chromosome. The $i^{th}$ holdout chromosome was identified and a data matrix, $A_{N \times (p+1)}$, was constructed by combining the binned genome from the remaining chromosomes ($1,2,\cdots, i-1,i+1,\cdots,21,22$), where $N=[n_{1} n_{2} \cdots n_{21} n_{22}]'$ and $n_{k}$ is the length of chromosome *k* after being binned into non-overlapping regions of resolution *r*, such that $k \ne i$. $p$, the number of annotations, together with the response vector **Y**, defined the column-wise dimension of the matrix $A$. Re-sampling was then performed on $A$, and an RF classifier was trained using 3-fold cross-validation to tune for the number of annotations to consider at each node (*mtry*). The number of trees (*ntree*) that were aggregated for each RF model was set to 500. The minimum number of observations per root node (*nodesize*) was set to .01% of the rows in the data. The binned data for the holdout chromosome $i$ was reserved for testing. Models were evaluated using Balanced Accuracy (BA), defined as the average of sensitivity and specificity:

@@

where TP refers to the number of bins correctly identified as containing a boundary (true positives), FP refers to the number of bins incorrectly identified as containing a boundary (false positives), TN refers to the number of bins correctly identified as not containing a boundary (true negatives), and FN refers to the number of bins incorrectly identified as not containing a boundary (false negatives). Each of these quantities is obtained from the confusion matrix created by validating the model on the test data. The process was repeated for each $i^{th}$ holdout chromosome, and performances were aggregated using mean and standard deviation.

@@

Many genomic annotations, notably architectural proteins, tend to exhibit an extensive pattern of colocalization (correlation) [@van2014insulator]. To suitably reduce the predictor space and improve computational efficiency, while maintaining optimal performance, we implemented recursive feature elimination (RFE). In doing so, we estimated the near-optimal number of necessary features to include in the best fitting RF model. The best performing subset was identified, ranging from 2 to the maximum number of features incremented by the power of 2. We then aggregated the predictive importance of the union of the optimal set of features across chromosomes using the mean decrease in node impurity among permuted features in out-of-bag samples to determine the most common and top-ranked annotations for predicting boundary regions.

@@

When evaluating performance across cell lines, a random forest model was built using ground truth boundaries and genomic annotations from one cell line such that the data matrix was constructed using genomic bins from $n-1$ chromosomes. The test data used to evaluate performance was constructed from the binned data of the $i^{th}$ holdout chromosome with genomic annotations from the other cell line. Thus, the models were trained on cell-line-specific data and tested on the other cell line. This process was repeated for each holdout chromosome. To evaluate performance, we constructed receiver operating characteristic (ROC) curves composed of the average sensitivities and specificities and reported the area under the curve (AUC).

@@

To investigate whether we could alleviate the resolution limitations of conventional domain calling tools, we developed *preciseTAD*. This algorithm leverages a random forest model in conjunction with density-based and partitioning techniques to predict boundaries at base resolution. A random forest model was built on the optimal combination of predictor type (predictor space), resampling technique (class balancing correction), and top-ranked annotations (most informative genomic elements) to maximally accurately predict boundaries at the _Hi-C resolution level_.

@@

We assessed the biological significance of our predicted boundaries by their association with several known molecular drivers of 3D chromatin architecture, including CTCF, RAD21, SMC3, and ZNF143 ChIP-Seq peaks. To do so, we used the *computeMatrix* and *plotHeatmap* tools offered by deepTools (version 2.0) [@ramirez2016deeptools] via integration with Galaxy (version 3.3.2.0.0) [@afgan2018galaxy]. Additionally, we compared the median $log_2$ genomic distances between TAD boundaries and the same top predictive ChIP-seq annotations using Wilcoxon Rank-Sum tests. Furthermore, we compared the overlap between predicted and called boundaries in GM12878 and K562 cell lines. Boundaries were first flanked by resolution, *r*, and overlaps were visualized using Venn diagrams from the `Vennerable` R package (version 3.1.0). Overlaps were further quantified using the Jaccard index defined as

@@

## Developing an ML framework for optimal TAD boundary prediction

We developed a machine learning (ML) framework aimed at determining the optimal set of data level characteristics necessary for precise prediction of boundaries of Topologically Associated Domains (TADs) and chromatin loops, collectively referred to as domain boundaries. We chose the random forest (RF) algorithm as our binary classification tool. The reason for it is two-fold: (1) to devise a tunable prediction rule in a supervised learning framework that is both robust to overfitting and able to handle multiple correlated predictors, and (2) to allow for an interpretable ranking of predictors [@boulesteix2012overview]. We use published Arrowhead-called TAD boundaries [@durand2016juicer] and Peakachu-predicted loop boundaries [@salameh2019supervised] as ground truth. Data from GM12878 and K565 cell lines at 5-100kb resolution (Arrowhead) and 10kb resolution (Peakachu) were used (Supplementary Table 1). We next developed *preciseTAD*, an R package that leverages an optimally built RF to precisely predict genomic coordinates of TAD and loop boundaries at base resolution from cell-type-specific genome annotation data.

Boundary regions were defined as genomic bins containing a called boundary ($Y = 1$), while non-boundary regions were defined as bins that did not contain a called boundary ($Y = 0$) (Figure 1A, see Methods). The total number of called TADs, their unique boundaries, and the number of genomic bins expectedly decreased with the decreased resolution of Hi-C data (Table 1, Supplementary Table 3). The number of non-boundary regions highly outnumbered boundary regions. Such a disproportional presence of examples in one class is known as a "class imbalance" problem that negatively affects predictive modeling [@prati2011survey; @wei2013role]. To address the class imbalance, we evaluated the effect of three resampling techniques. *Random over-sampling* (ROS) was defined as sampling with replacement from the minority class (boundary regions). *Random under-sampling* (RUS) was defined as sampling with replacement from the majority class (non-boundary regions). Lastly, we tested *Synthetic minority over-sampling technique* (SMOTE), which is a combination of both random over- and under-sampling to create balanced classes [@chawla2002smote] (see Methods).

A total of 77 cell-line-specific genomic annotations were used to build the predictor space. These included BroadHMM chromatin segmentation, histone modifications (HM), and transcription factor binding sites (TFBS) (Supplementary Table 2). Four feature engineering procedures were developed to quantify the association between genomic annotations and bins (Figure 1B). These included signal strength association (Signal), direct (OC), proportional (OP), and spatial ($log_2 + 1$ Distance) relationships (see Methods, Supplementary Figure 1).

![**Resolution-specific data construction and feature engineering used for downstream random forest modeling.** (A) The linear genome was binned into non-overlapping resolution-specific intervals (see Methods). The response vector **Y** was defined as 1/0 if a genomic bin overlapped/did not overlap with a TAD boundary. (B) Four types of associations between bins and genomic annotations (green shapes) were considered to build the predictor space including Average Peak Signal (Signal), Overlap Counts (OC), Overlap Percent (OP), and $log_{2}$ distance (Distance).](Figures/figure1.png)

In total, we considered combinations of data from two cell lines *L = {GM12878, K562}*, five resolution *R = {5 kb, 10 kb, 25 kb, 50 kb, 100 kb}*, four types of predictor space *P = {Signal, OC, OP, Distance}*, and three re-sampling technique *S = {None, RUS, ROS, SMOTE}* (Figure 2). Once the model inputs were established, a random forest classifier was trained on $n-1$ autosomal chromosomes, while reserving the $i^{th}$ chromosome for testing. Three-fold cross-validation was used to tune the *mtry* hyperparameter, while *ntree* and *nodesize* were fixed to default values of 500 and 1, respectively. Model performance was evaluated by aggregating the mean balanced accuracy (BA) across each holdout chromosome (see Methods). These strategies allowed us to select the best performing model in an unbiased manner.

![**A diagram of the machine learning framework used to build TAD/loop boundary region prediction models.** Step 1 defines the dimensions of the data matrix $A_{N \times (p+1)}$ used for training and testing by establishing a resolution, $r$, at which to bin the linear genome into $N$ genomic bins for each chromosome, except a holdout chromosome *i*. The response vector **Y** is defined as a boundary region ($Y=1$) if it overlaps with a genomic bin (else $Y=0$). A feature engineering procedure is selected to define the feature space of $A$ composed of $p$ genomic annotations. Step 2 is to reserve the predictor-response matrix for the holdout chromosome *i* as the test data. Step 3 is to apply a resampling technique to the training data to address the class imbalance. Step 4 is to train the random forest model and perform 3-fold cross-validation to tune for the mtry parameter. Finally, step 5 is to validate the model on the separate test data composed of the binned data from the holdout chromosome *i* and to evaluate model performance using balanced accuracy (BA).](Figures/figure2.png)

@@

Expectedly, when using data with class imbalance present, that is, no resampling, the models exhibited low balanced accuracies, with minimal variability among different resolutions (Figure 3). Similarly, poor performances were found when using ROS. However, RUS and SMOTE re-sampling led to a drastic improvement in performance, especially at higher resolutions. We found that RUS marginally outperformed SMOTE as the optimal class balancing technique for all resolutions and predictor types when predicting TAD boundary regions.

Additionally, we found that using a distance-type predictor space yielded substantially higher balanced accuracy compared to the peak signal strength, overlap count, and overlap percent predictor types. As with class balancing techniques, this improvement was less evident at lower resolutions, with results consistent for K562 (Supplementary Figure 2). Random forest models built on Peakachu-defined loop boundaries with SMOTE resampling only marginally outperformed RUS for both GM12878 and K562 cell lines (Supplementary Figure 3). Furthermore, 5 kb resolution genomic bins led to the optimal prediction for both TAD and loop boundaries, on both cell lines. Our results indicate that random under-sampling, distance-type predictors, and high-resolution Hi-C data provide the optimal set of data level characteristics for both TAD and chromatin loop boundary prediction.

![**Distance-type predictors combined with random under-sampling at 5 kb resolution genomic bins yielded optimal TAD boundary region predictive performance for GM12878 Arrowhead TAD boundaries.** Averaged balanced accuracies are compared across resolution, within each predictor-type: (A) Signal, (B) OC, (C) OP, and (D) Distance, and across resampling techniques: no resampling (None; red), random over-sampling (ROS; green), random under-sampling (RUS; blue), and synthetic minority over-sampling (SMOTE; purple). Error bars indicate standard deviation away from the mean performance across each holdout chromosome used for testing.](Figures/figure3.png)

@@

We hypothesized that the type of genomic annotations may also affect predictive performance. Using the established optimal settings (RUS, Distance, 5 kb bins), we built separate random forest models using BroadHMM chromatin segmentation, histone modifications (HM), and transcription factor binding sites (TFBS). We found that models built on TFBSs outperformed other annotation-specific models, with results consistent for loop boundaries, on both cell lines (Figure 4A; Supplementary Figures 4A). These results suggest that TFBS are the primary drivers of TAD and loop boundary formation in both GM12878 and K562.

## Predictive importances confirmed the biological role of CTCF, RAD21, SMC3, and ZNF143 for boundary formation

We sought to further optimize our boundary region prediction models by recursive feature elimination to avoid overfitting and selecting only the most influential TFBS across all autosomal chromosomes. We were able to obtain near-optimal performance using approximately eight TFBS (Figure 4B; Supplementary Figures 4B). However, given that we trained our model on chromosome-specific data, the most significant annotations varied for each chromosome set. To determine transcription factors most important for boundary prediction, we clustered the predictive importance (mean decrease in accuracy) of the top eight significant TFs across chromosomes. We found that similar to other studies [@sanborn2015chromatin; @weintraub2017yy1; @gonzalez2017znf143], CTCF, RAD21, SMC3, and ZNF143 transcription factors were consistently predictive of TAD and loop boundaries (Figure 4C; Supplementary Figures 4C). We optimized our model by only considering these top informative TFBS when building the random forest, thereby decreasing computational burden while maintaining high predictive performance. In summary, our model was able to yield the known molecular drivers of the loop extrusion model [@hansen2018recent;@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self; @Mirny:2019aa; @Davidson:2019aa].
